Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour

### 2018-02-07 Cars/paper.tex
more paper changes mostly

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\section{Introduction}

\begin{itemize}
\item Self driving Cars are becoming more common, eg Waymo, Tesla, GM.
\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg. Cameras and LiDAR), such as their position, velocity, and steering angle (Woo et al. 2016a; Tian, Pei, Jana, and Ray 2017).
\item The Self Driving Car forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
\item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al. 201b6; Tomar, Verma, and Tomar 2014; Woo et al. 2017; Geng et al. 2017).
\item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
\item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can be very different from driver to driver (Chen 2017).
\item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
\item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
\item The self driving car is constantly observing additional information about the vehicles around it. Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
\item A three step strategy is employed to forecast the future trajectory of newly observed vehicles:
\begin{enumerate}
\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution. This can occur before the self driving car is on the road.
\item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
\item Update the parameter vector and all forecasts at short intervals.
\end{enumerate}
\item This paper introduces a Variational Bayes updating method for the third step, where the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
\item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
\item This is important for two reasons:
\begin{enumerate}
\item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
\item Self driving cars take in up to 1GB of data a second (several websites report this as quotes from different people eg. Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
\end{enumerate}
\item Compare Bayes MAP point estimates to a neural network and constant acceleration / angle model.
\item Error Metric: Euclidean Distance from predicted position to actual position
\item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB. I am also using MCMC, so we can compare either VB method to this, theoretically MCMC > VB Standard > VB Updating
\end{itemize}

@@

\begin{itemize}
\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes ~5000 cars for a 2000 feet section of the US 101 Highway.
\item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
\item Around 2000 of these cars changed lanes or stopped moving somewhere in the sample and are discarded.
\item There was some curvature in the road which means there are systematic movements in the car paths.
\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010), instead of relative to the Earth (ie. ignores curves)
\item Prior research fits a polynomial curve to detected lane markings to build a model of the road lane edges (Woo et. al. 2016) 
\item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint. Relative co-ordinates are found separately for each lane using the movements of 100 cars / lane by the following procedure:
\begin{enumerate}
\item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
\item Estimate independent smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
\item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
\item Define new coordinates 
\begin{align}
x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right) - \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t} \label{yRel}
\end{align}
\end{enumerate}
\item (\ref{xRel}) \textit{really} needs a diagram.
\item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
\end{itemize}

@@

The position of any vehicle at time $t$ can be fully determined by its initial position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the vehicle's velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
\begin{align}
x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}). \label{yEq},
\end{align}
where $\delta_{t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{t} = x^*_{t-1}$.
From this relationship, the driver's key inputs to motion can be extracted by
\begin{align}
\delta_{t} &= \tan^{-1}\left(\frac{(y^*_{t} - y^*_{t-1})}{(x^*_{t} - x^*_{t-1})} \right) \label{dEq} \\
v_{t} &= \sqrt{(x^*_{t} - x^*_{t-1})^2 + (y^*_{t} - y^*_{t-1})^2} \label{vEq} \\
a_{t} &= v_{t} - v_{t-1}. \label{aEq}
\end{align}
where $a_{t}$ denotes the acceleration at time $t$.

@@

Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity within each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model. In this case each $\theta_i$ is treated as an independent draw from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that 
\begin{equation}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j).
\end{equation}
For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$. $k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with fixed hyperparameters $\alpha_1, \alpha_2, \dots, \alpha_K$. Defining $\beta = \{\mu_i, \Sigma_i, i = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.

@@

Variational Bayes (Jordan et al. 1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
\begin{equation}
\label{KL-div}
KL(q(\theta) || p(\theta | z) = \int_{\theta} q(\theta) \left(\log(q(\theta)) - \log(p(\theta | z))\right) d\theta.
\end{equation}
As $p(\theta | z)$ is typically intractable, VB maximises $\mathcal{L}(q)$ where
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
\end{equation}
which is equivalent to minimising (\ref{KL-div}), where $\lambda$ is the parameter vector of $q(\theta)$.
This is done with stochastic gradient ascent, taking Monte Carlo estimates of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$ (Hoffman et al. 2013)
and repeating updates of the form
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} 
\end{equation}
which is guaranteed to converge to a local maximum (Robbins and Munro, 1951) if $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}
In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of Kingma and Ba (2015).

There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the Score Estimator of Ranganath, Gerrish and Blei (2014), 
\begin{equation}
\label{scoreDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_i))}{\delta \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
\end{equation}
and the Reparameterised Estimator of Kingma Welling (2014),
\begin{equation}
\label{rpDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^M \frac{\delta f(\lambda, \epsilon_i)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_i)} + \frac{\delta J(\lambda, \epsilon_i)}{\delta \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
\end{equation}
where $r(\epsilon)$ is some distribution without any free parameters (eg. Uniform, Standard Normal), $f(\lambda, \epsilon)$ is some differentiable function that transforms $\epsilon$ to $\theta$, and $J(\lambda, \epsilon)$ is the Jacobian Matrix of this function. (\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the derivative estimator (\ref{scoreDeriv}) must be used. To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$. Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.

@@

\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the Score Estimator for updating Variational Bayes, 

@@

\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the Reparameterised Estimator for updating Variational Bayes,

@@

\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)

@@

\section{Empirical}

\begin{figure}[ht]
\centering
\includegraphics[width = 0.8\textwidth]{HierSingleKDE.png}
\caption{Top: A Kernel Density Estimate of the distribution of posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ modelled independently with no hierarchical structure. Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities. The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
\label{fig:HierSingleKDE}
\end{figure}

Models:
\begin{enumerate}
\item No prior information for the new car - very flat priors.
\item Prior information from hierarchical model with one component
\item Prior information from large mixture hierarchical model
\end{enumerate}
\begin{itemize}
\item In each case the VB posterior approximation $q(\theta)$ has the same family as the prior $p(\theta)$, so either a multivariate normal or a mixture of multivariate normals. This is so we can replace the prior with the VB fit at each step of the updating method easier. 
\item The Hierarchical model is fit for 2000 cars, leaving 873 for forecast evaluation.
\end{itemize}

Methods:
\begin{enumerate}
\item MCMC
\item Regular VB (fit each update to the new sample)
\item Updating VB (fit to new data using the previous VB fit as a prior)
\end{enumerate}

\begin{center}
\begin{table}[h]
\resizebox{\textwidth}{!}{%
\begin{tabular}{| l | c | c | c |}
\hline
& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
\hline
MCMC & \multicolumn{3}{c|}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
& \multicolumn{3}{c|}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
\hline
VB  & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpDeriv}) to update parameters of a} & Variational Bayes using (\ref{scoreDeriv}) to update parameters of a six \\
Standard & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\
\hline
VB & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpUpdate}) to update parameters of a} &  Variational Bayes using (\ref{scoreUpdate}) to update parameters of a six \\
Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ 
\hline
\end{tabular}}
\label{tableAlg}
\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
\end{table}
\end{center}

There are six naive forecast models with constant estimates of the future values of $v_t$ and $\delta_t$ for $t = T+1, \dots, T+30$. $v_t$ is estimated by either the current velocity, $v_T$ or the average of the previous ten lags of $v$, $1/10 \sum_{s = T - 9}^T v_s$. $\delta_t$ is estimated by one of the current steering angle, $\delta_T$, the average of the previous ten lags of $\delta$, $1/10 \sum_{s = T - 9}^T \delta_s$ or by $\pi/2$, i.e. the car is moving directly ahead. Table \ref{tableNaive} classifies each naive model based on the combination of velocity and angle estimates.
\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|}
\hline
& $\hat{\delta}_t = \delta_T$ & $\hat{\delta}_t = 1/10 \sum_{s=T-9}^T \delta_s$ & $\hat{\delta}_t = \pi/2$ \\
\hline
$\hat{v}_t = v_T$ & Naive 1 & Naive 2 & Naive 3 \\
$\hat{v}_t = 1/10\sum_{s=T-9}^T$ & Naive 4 & Naive 5 & Naive 6\\
\hline
\end{tabular}
\end{center}
\label{tableNaive}
\caption{Classification of naive forecast models by future velocity and steering angle estimators}
\end{table}

>>>>

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\title{Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour}
\begin{document}
\maketitle
\section{Introduction}
\item Self driving Cars are becoming more common, eg Waymo, Tesla, GM.
\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
\item While Self Driving Cars reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg. Cameras and LiDAR), such as their position, velocity, and steering angle (Woo et al. 2016a; Tian, Pei, Jana, and Ray 2017).
\item The Self Driving Car forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
\item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al. 201b6; Ding et al. 2012; Woo et al. 2017; Geng et al. 2017).
\item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
\item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can depend both on the be very different from driver to driver (Chen 2017) which may depend on a particular vehicle's position relative to surrounding traffic.
\item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
\item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
\item The self driving car is constantly observing additional information about the vehicles around it. Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
\item A three step strategy is employed to forecast the future trajectory of newly observed vehicles:
\begin{enumerate}
\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution. This can occur before the self driving car is on the road.
\item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
\item Update the parameter vector and all forecasts at short intervals.
\end{enumerate}
\item This paper introduces a Variational Bayes updating method for the third step, where the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
\item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
\item This is important for two reasons:
\begin{enumerate}
\item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
\item Self driving cars take in up to 1GB of data a second (several websites report this as quotes from different people eg. Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
\end{enumerate}
\item Compare Bayes MAP point estimates to a neural network and constant acceleration / angle model.
\item Error Metric: Euclidean Distance from predicted position to actual position
\item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB. I am also using MCMC, so we can compare either VB method to this, theoretically MCMC $\geq$ VB Standard $\geq$ VB Updating

@@

\begin{itemize}
\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes cars for a 2000 feet section of the US 101 Highway.
\item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
\item Around 2000 of these cars changed lanes or stopped moving somewhere in the sample and are discarded.
\item There was some curvature in the road which means there are systematic movements in the car paths.
\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010), instead of relative to the Earth (ie. ignores curves)
\item Prior research fits a polynomial curve to detected lane markings to build a model of the road lane edges (Woo et. al. 2016) 
\item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint. Relative co-ordinates are found separately for each lane using the movements of 100 cars / lane by the following procedure:
\begin{enumerate}
\item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
\item Estimate independent smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
\item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
\item Define new coordinates 
\begin{align}
x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right) - \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t} \label{yRel}
\end{align}
\end{enumerate}
\item (\ref{xRel}) \textit{really} needs a diagram.
\item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
\end{itemize}

@@

The position of any vehicle at time $t$ can be fully determined by its initial position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the vehicle's velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
\begin{align}
x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}) \label{yEq},
\end{align}
where $\delta_{t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{t} = x^*_{t-1}$.
From this relationship, the driver's key inputs to motion can be extracted by
\begin{align}
\delta_{t} &= \tan^{-1}\left(\frac{(y^*_{t} - y^*_{t-1})}{(x^*_{t} - x^*_{t-1})} \right) \label{dEq} \\
v_{t} &= \sqrt{(x^*_{t} - x^*_{t-1})^2 + (y^*_{t} - y^*_{t-1})^2} \label{vEq} \\
a_{t} &= v_{t} - v_{t-1}. \label{aEq}
\end{align}
where $a_{t}$ denotes the acceleration at time $t$.

@@

Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity within each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model. In this case each $\theta_i$ is treated as an independent draw from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that 
\begin{equation}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j).
\end{equation}
For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$. $k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi} = \{\pi_{1}, \dots, \pi_{K}\}$, where $\boldsymbol{\pi}$ has a Dirichlet prior with fixed hyperparameters $\alpha = \alpha_j, j = 1 , \dots, K$. Defining $\beta = \{\mu_j, \Sigma_j, \pi_j, j = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.

\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
\caption{Top: A Kernel Density Estimate applied to the individual posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$, modelled independently without any hierarchical structure. Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities. The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
\label{fig:HierSingleKDE}
\end{figure}

@@

Variational Bayes (VB, Jordan et al. 1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
\begin{equation}
\label{KL-div}
KL(q(\theta) || p(\theta | z) = \int_{\theta} q(\theta) \left(\log(q(\theta)) - \log(p(\theta | z))\right) d\theta.
\end{equation}
As $p(\theta | z)$ is typically intractable, VB maximises $\mathcal{L}(q)$ where
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
\end{equation}
which is equivalent to minimising (\ref{KL-div}), where $\lambda$ is the parameter vector of $q(\theta)$.
This is done with stochastic gradient ascent, taking Monte Carlo estimates of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$ (Hoffman et al. 2013)
and repeating updates of the form
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} 
\end{equation}
which is guaranteed to converge to a local maximum (Robbins and Munro, 1951) if $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}
In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of Kingma and Ba (2015).

There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of Ranganath, Gerrish and Blei (2014), 
\begin{equation}
\label{scoreDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_i))}{\delta \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
\end{equation}
and the reparameterised estimator of Kingma and Welling (2014),
\begin{equation}
\label{rpDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^M \frac{\delta f(\lambda, \epsilon_i)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_i)} + \frac{\delta J(\lambda, \epsilon_i)}{\delta \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
\end{equation}
where $r(\epsilon)$ is some distribution without any free parameters (eg. Uniform, Standard Normal), $f(\lambda, \epsilon)$ is some differentiable function that transforms $\epsilon$ to $\theta$, and $J(\lambda, \epsilon)$ is the Jacobian Matrix of this function. (\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the score estimator (\ref{scoreDeriv}) must be used. To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$. Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.

@@

\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the score estimator for updating Variational Bayes, 

@@

\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the reparameterised estimator for updating Variational Bayes,

@@

\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda_T} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)

@@

\section{Forecast Evaluation}

\begin{itemize}
\item 4071 vehicles are available in the sample
\item 500 are used to estimate splines and straighten the road to relative coordinates as described in Section 2
\item 2000 are used to train any models used
\item 1571 are available for testing
\item Each second (10 observations) models are fit to the sample of $z_{N+1, 1:T}$ and forecasts of $z_{N+1, T+1:T+30}$ are made, corresponding to a three second forecast horizon
\item Forecasts are compared in terms of predictive log-score (where a predictive density is available) and Euclidean error.
\end{itemize}

\subsection{Heterogenous Time Series Models}

There are three versions of the heterogenous time series model described in Section 3, differentiated by the form of the prior, $p(\theta_{N+1} | z_{1:N})$.
\begin{enumerate}
\item Non-informative prior for the new car, where $p(\theta | z_{1:N})$ is multivariate normal with mean vector $(-5, -5, 0, 0, 0, 0)$ and diagonal variance matrix $10 \mathbb{I}$.
\item Informative prior where $p(\theta | z_{1:N})$ is multivariate normal with a mean vector and variance matrix estimated by the posterior mean of $\mu$ and $\Sigma$ in a hierarchical model with $K = 1$ fit by MCMC to the 2000 training vehicles. 
\item Informative prior where $p(\theta | z_{1:N})$ is a six component mixture of multivariate normals with mean vectors, variance matrices and mixture weights estimated by the posterior means of each $\mu_j$ and $\Sigma_j$ and $\pi_j$ in a hierarchical model with $K = 6$ fit by MCMC to the 2000 training vehicles. 
\end{enumerate}
Each model is fit by each of Metropolis-Hasting MCMC, standard Variational Bayes where the entire observed history of $z_{N+1, 1:T}$ is used for each model fit, and the updating Variational Bayes process described by Section 4.2 where only the most recent observed data $z_{N+1, S+1:T}$ is used in each iteration. In each Variational Bayes algorithm the approximating distribution has the same functional form as the prior used, a multivariate normal distribution for the non-informative prior and $K = 1$ hierarchy, and a six component mixture of multivariate normals for the $K = 6$ hierarchy. Details of each case are provided in Table \ref{tableAlg}.
\begin{center}
\begin{table}[h]
\resizebox{\textwidth}{!}{%
\begin{tabular}{| l | c | c | c |}
\hline
& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
\hline
MCMC & \multicolumn{3}{c|}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
& \multicolumn{3}{c|}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
\hline
VB  & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpDeriv}) to update parameters of a} & Variational Bayes using (\ref{scoreDeriv}) to update parameters of a six \\
Standard & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\
\hline
VB & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpUpdate}) to update parameters of a} &  Variational Bayes using (\ref{scoreUpdate}) to update parameters of a six \\
Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ 
\hline
\end{tabular}}
\label{tableAlg}
\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
\end{table}
\end{center}

\subsection{Competing Models}

\subsubsection{Constant Angle and Acceleration Models}
